Astrophysical calibration of gravitational-wave detectors

### 2015-04-09 docs/cosmic_cal.tex
A bit more tidying up.

% Describe what we are going to do in this section

% Describe the signal model (Taylor F2 etc...)

% Define the detectors and the noise epoch we will be using

% Define the likelihood function we will be using

% describe what parameters we will be considering

% describe the priors

% mention the no-noise issue

@@

In this project the hypothesis considered, $\curlH$, simply states that the
data, $\tilde{d}(f),$ collected by a generic detector contains the GW signal,
$\gws(f)$, and some noise, $\tilde{n}(f)$~\cite{JVei},

\begin{equation}
  \label{eq:data}
  \tilde{d}^D(f) = \gws^D(f) + \tilde{n}^D(f)
\end{equation}

where $D$ represents the detector site. In this project there will be three
detectors in question: LIGO Livingston (L), LIGO Hanford (H), and Virgo (V); $D
= \{L,H,V\}$. Testing different hypotheses is the method used to determine
whether a data set contains a GW signal or not~\cite{SathSchutz}. This was not
explored in this project so the above hypothesis is consistent throughout this
paper.

In this paper, a brief overview of the mathematical breakdown for the waveform
is shown. For a more rigorous derivation, see~\cite{SathSchutz} and references
there in. The waveform used throughout this project is a simplified system
where the neutron stars are non rotating about their axes, in a perfectly
circular inspiralling orbit; known as the TaylorF2 model~\cite{TF2,pN}. A more
complex and realistic model could be tested in future for this calibration
method. Parameter estimation using a variety of complex waveforms are carried
out in~\cite{peadvanced}.

The general equation for a GW signal detected at a generic detector, in the
reference frame of the detector site, is given as~\cite{SathSchutz}

\begin{equation}
  \label{eq:gravsig}
   \gws^D(f) = \big[ F_{+}^D(\alpha, \delta, \psi)\gws_{+}(f) +
F_{\times}^D(\alpha, \delta, \psi)\gws_{\times}(f)\big]
\end{equation}

where $F_{+,\times}(\alpha, \delta, \psi)$ are the antenna response function of
the detector, dependent on the right ascension, $\alpha$, declination,
$\delta$, the polarisation angle $\psi$, and time. Testing the antenna patterns
is carried out in section~\ref{sec:rndsp} of this report when the sky
positions, $(\alpha, \delta)$, are randomised across the entire field of view
for the detector.

The polarisation amplitudes of the GW, given by $\gws_{+}(f)$, and
$\gws_{\times}(f)$, represent the direction with which the GW compresses and
stretches space-time. The form of these two amplitudes depends on the source
being analysed. As mentioned above, the model used throughout this project was
the TaylorF2 model. The parameters defining the GW polarisation amplitudes
using the TaylorF2 model are:

\begin{itemize}
\item the total mass, $M = m_{1}+m_{2}$, where $m_{1,2}$ are the masses of the
binary stars
\item the symmetric mass ratio $\eta = \frac{m_{1}m_{2}}{(m_{1}+m_{2})}$
\item chirp mass, $\mathcal{M} = \eta^{3/5}M$
\item luminosity distance, $D_{L}$, measured in Mpc ($1 Mpc \approx
3.1\times10^{22}m  $)
\item the angle of inclination $\iota$. This is the angle from the source,
taken as the angle between the normal of the orbital plane and the observer
orientation.
\item phase of the wave, $\psi$, which defines the polarisation of the angle of
the GW, shown in equation~\ref{eq:psi}.
\item the coalescence time, $t_c$, and phase of coalescence, $\phi_c$.
\end{itemize}

Together with the sky position, these parameters form the vector $\vec{\Theta}
= \{\mathcal{M}, \eta, D_{L}, \iota, \alpha, \delta, \psi, t_{c}, \phi_{c}\}$
which defines the entire GW signal.  The form of the polarisation amplitudes
using the TaylorF2 model are given as~\cite{SathSchutz,pN,TF2}

\begin{equation}
  \gws_{+}(f) = \frac{1+\cos^{2}\iota}{2D_{L}}
\frac{\mathcal{M}^{5/6}}{\pi^{2/3}} \sqrt{\frac{5}{24}}f^{-7/6}e^{-i\psi(f)}
\end{equation}
\begin{equation}
  \gws_{\times}(f) =
i\frac{\cos^{2}\iota}{D_{L}}\frac{\mathcal{M}^{5/6}}{\pi^{2/3}}
\sqrt{\frac{5}{24}}f^{-7/6}e^{-i\psi(f)}
\end{equation}

where the phase of the GW, $\psi$, is
\begin{equation}
  \label{eq:psi}
  \psi(f) = 2\pi f t_c - \phi_c + \psi_N(\eta)\sum\limits_{k=0}^4{\psi_k(\eta)
(\pi(m_1 + m_2)f)^{(k-5)/3}}.
\end{equation}

The terms $\psi_N$ and $\psi_k$ are the standard Newtonian and post-Newtonian
coefficients. Derivations of these two equations can be found in~\cite{pN} and
references there-in. The simulated gravitational wave signal used throughout
this project can be seen in figure~\ref{fig:signal}. As one can see, the signal
strength is very faint (an order of magnitude around $10^{-24}$). This is the
the biggest problem in detecting GWs.

As mentioned above, the source considered throughout this project thought to
produce EM radiation; SGRBs. In this project, we assume that SGRBs of the GW
source described above have been observed. If a GRB has been seen, then the
redshift of the merger can be calculated. This then allows for the luminosity
distance to be calculated and in turn other defining parameters can be obtained
(or known to a degree of uncertainty)~\cite{SathSchutz}.

%More explination
Finally, the last term in equation~\ref{eq:data} must be discussed; the noise
of the model, $\tilde{n}(f)$. Throughout this project, we have added no noise
to the signal, although as discussed later we do include a noise estimate in
the likelihood.

The prior probability function, $p(\vec{\theta} | \curlH)$ in
equation~\ref{eq:bayesgw}, represents our prior knowledge in $\theta$ before
the analysis.The range of parameters used throughout this paper were chosen due
to their inability to be justifiably known through a GRB assumption. The
parameters were drawn from the ranges for any signals generated in this
analysis, given in table 1.

\begin{table}
  \centering
  \begin{tabular}{|c|c|}
%  \caption{Prior parameters}
    \hline
    Parameter & Distribution \\
    \hline
    $\psi$   &  U($0,\pi/2$)\\
    $\iota$   & N($0,0.1$) \\
    $\phi_c$  & U($0,\pi/2$)\\
    $t_c$     & U($t_{o}-0.01, t_{o}+0.01$) \\
    $s$       & U(0,2) \\
    \hline
  \end{tabular}
  \label{tab:prior}
  \caption{\textit{The parameters used throughout the analysis with some degree
of uncertainty. U(a,b) is a uniform distribution between a and b. N($\mu,
\sigma$) is a normal distribution, with a mean $\mu$ and standard deviation
$\sigma$. $t_{o} = 9\times10^{8}$}}
\end{table}

Through the assumption made above allows us to constrain the system,
maintaining a small number of parameters in the prior. For a realistic
parameter estimation of the source, a larger range of values with a bigger
distribution. This was not tested due to time constraints but is proposed as
work for the future.

The likelihood functiona used throughout this project is given for a generic
detector D as~\cite{JVei},

\begin{equation}
  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp(-(d^{D}(f) -
h^{D}(f)|d^{D}(f) - h^{D}(f))/2)
\end{equation}

where $(\cdot |\cdot )$ is a representation of the inner product. This
derivation is given in the appendix. Using the inner product of two real
functions, and approximating for discrete integer values, the likelihood then
becomes

\begin{equation}
  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp\Bigg[ 2\Delta
f\sum_{k>0}\frac{|\tilde{d}(f_{k}) - \tilde{h}(f_{k})|^{2}}{S(f_{k})} \Bigg]
\end{equation}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{PSD}
  \caption{\textit{Noise spectral density for advanced LIGO and Virgo.}}
  \label{fig:psd}
\end{figure}

where $S(f_{k})$ is the \textit{power spectral density} (PSD). This defines the
sensitivity of the detector as it a measure of the strain in the detector due
to noise~\cite{SathSchutz,sens}. The PSD for advanced LIGO and Virgo are shown
in figure~\ref{fig:psd}. The figure shows that the strain effect for advanced
LIGO is less than it is for Virgo; LIGO has a better sensitivity. This is a
factor in analysing the calibration error, given in section~\ref{sec:results}.
The subscript $k$ represents discrete values of relevant frequency bins (As
carrying out the integral would become computationally costly). This likelihood
will be used throughout the project. The factor of $\Delta f$ comes from the
approximation of the inner product of two real functions (as it is integrated
over frequency space).

The aim of this project was to determine how well the calibration error could
be parameterised in a software injected gravitational wave source. The error
was estimated through numerous situations but the basis of the simulations
remained the same throughout the project: a single source of GWs at varying
signal to noise ratios. The GRB assumption was analysed as well by relaxing
some of the defining GW paramaters and adding some uncertainty into their
value, given in section~\ref{sec:pandl}.

These values where chosen due to the assumed counterpart EM observation. These
values could not be justified by that assumption so they were added into the
prior with reasonable uncertainty.

The estimation of the error, given in its simplest form as a scale factor, was
performed for numerous systems: single detector, a network of detectors, random
sky positions, etc. Before the parameter estimation was carried out, the
relationship between distance of the source relative to the detector and the
signal to noise ratio (SNR) needed to be found.

The SNR is given by
\begin{equation}
  \label{eq:snr}
  SNR = \sqrt{\Bigg( \frac{2 |\gws^D(f)|^2}{S^D(f)} df \Bigg)}
\end{equation}

>>>>

% Describe what we are going to do in this section
Let us consider a single \ac{GW} detection of the inspiral stage of a \ac{BNS}
merger and assume that this signal has been associated with a potential
\ac{sGRB} detection. We therefore also assume that the inclination angle
of the binary is consistent with a given \ac{sGRB} beaming angle.  Finally we
assume that a host galaxy has been identified for the source and that the sky
position is therefore known to high precision.

% Describe the signal model (Taylor F2 etc...)
We use the post-Newtonian Taylor F2 waveform approximation for modelling the
signal whereby the \ac{GW} phase for 2 point-particles of mass $m_{1}$ and
$m_{2}$ is given by
%
\begin{equation}
\label{eq:gwphase}
  \Psi(f) = 2\pi f t_\mathrm{c} - \phi_{\mathrm{c}} - \frac{\pi}{4} +
\frac{3}{128\eta\nu^{5}}\sum\limits_{k=0}^{7}\alpha_{k}\nu^{k}.
\end{equation}
%
Here $\eta=m_{1}m_{2}/M^{2}$ is the symmetric mass-ratio with $M=m_{1}+m_{2}$
the total mass.  The post-Newtonian expansion parameter is $\nu=(\pi M
f)^{1/3}$ and the coefficients are $\alpha$.

The frequency-domain polarisation amplitudes are then defined as
%
\begin{eqnarray}
  \gws_{+}(f) &=& \frac{1+\cos^{2}\iota}{2D_{L}}
\frac{\mathcal{M}^{5/6}}{\pi^{2/3}}
\sqrt{\frac{5}{24}}f^{-7/6}e^{-i\Psi(f)}\\
  \gws_{\times}(f) &=&
 i\frac{\cos^{2}\iota}{D_{L}}\frac{\mathcal{M}^{5/6}}{\pi^{2/3}}
\sqrt{\frac{5}{24}}f^{-7/6}e^{-i\Psi(f)}
\end{eqnarray}
%
where the chirp-mass $\mathcal{M}$ is defined as $\mathcal{M}=M\eta^{3/5}$,
and $\iota$ is the inclination angle.  The \ac{GW} strain measured
at the $k$'th detector is then given by
%
\begin{equation}
  \label{eq:gravsig}
   \gws^{k}(f) = \big[ F_{+}^{k}(\alpha, \delta, \psi)\gws_{+}(f) +
F_{\times}^D(\alpha, \delta, \psi)\gws_{\times}(f)\big]
\end{equation}
%
where $F_{+},F_{\times}$ are the antenna response functions which are dependent
upon the polarisation angle $\psi$ and the sky position of the source
$\alpha,\delta$.

% Define the detectors and the noise epoch we will be using
For this analysis we will consider advanced generation instruments operating
at their design sensitivity.  The network we consider is the 3 detector
H1--L1--V1 configuration.

% Define the likelihood function we will be using
We consider that the measured data in any detector is defined as
%
\begin{equation}
  d^{k}(f)=\tilde{n}^{k}(f)+s\gws_{k}(f)
\end{equation}
%
where $\tilde{n}(f)$ is the Fourier transform of Gaussian noise drawn from a
distribution consistent with the power-spectral-density of the advanced
detector noise.  The likelihood function for application to discretely sampled
data is 
%
\begin{equation}
  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp\Bigg[ 2\Delta
f\sum_{k>0}\frac{|\tilde{d}(f_{k}) - \tilde{h}(f_{k})|^{2}}{S(f_{k})}\Bigg]
\end{equation}
% 
where $\Delta f$ is the frequency resolution of the frequency domain data.

% describe what parameters we will be considering and their priors
We assume uniform prior distributions on the unknown nuisance parameters
$\phi_{\mathrm{c}},\psi$ on their standard ranges. The time of coalescence is
restricted to a prior range of $\pm 10$ msec around the true value.  Given our
assumption of an \ac{sGRB} counterpart and host galaxy, we treat the sky
position as exactly known. We also consider the cosine of the inclination angle to be drawn
from a zero-mean Gaussian with standard deviation $0.1$ consistent with an
uncertainty on \ac{sGRB} opening of XXX~\cm{What was exactly done here?}.  The
main parameter of interest, the calibration scale factor $s$ is assumed to have
a uniform prior distribution on the range $(0,2)$.

% mention the no-noise issue
We note that the analysis performed here used zero-noise realisations in which
no actual noise was added to our signals.  The likelihood functions still
contain the correct \ac{PSD} and hence the accuracy with which the parameters
are recovered (indicated by the widths of the posterior distributions) is
representative of the statistical errors in realistic noise scenarios.

@@
